---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crann

<!-- badges: start -->
[![pkgdown](https://github.com/christopherkenny/crann/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/christopherkenny/crann/actions/workflows/pkgdown.yaml)
<!-- badges: end -->

`crann` (Irish for "tree") implements spanning tree algorithms for undirected graphs represented as [`adj`](https://christophertkenny.github.io/adj/) adjacency lists.
It provides tools for counting, enumerating, sampling, and analyzing the spanning trees of a graph.

## Installation

You can install the development version of `crann` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("christopherkenny/crann")
```

## Examples

```{r setup, message = FALSE, warning = FALSE}
library(crann)
library(adj)
```

We build a complete graph on four vertices (K4) to use throughout these examples.

```{r graph}
k4 <- adj(list(c(2L, 3L, 4L), c(1L, 3L, 4L), c(1L, 2L, 4L), c(1L, 2L, 3L)))
k4
```

### Counting spanning trees

`count_spanning_trees()` uses Kirchhoff's matrix tree theorem to count spanning trees exactly.
K4 has 16.

```{r count}
count_spanning_trees(k4)
```

### Enumerating spanning trees

`enumerate_spanning_trees()` returns all spanning trees as a list of `adj` objects using Winter's (1986) contraction-based algorithm.

```{r enumerate}
trees <- enumerate_spanning_trees(k4)
length(trees)
trees[[1]]
```

For downstream computation, `enumerate_spanning_trees_edges()` returns the same trees as a compact integer matrix with n−1 rows and 2t columns.
Columns 2k−1 and 2k hold the edge endpoints of spanning tree k.
This avoids per-tree memory allocation and is substantially faster for graphs with many spanning trees.

```{r enumerate-edges}
mat <- enumerate_spanning_trees_edges(k4)
dim(mat)  # (n-1) x (2 * n_trees)
mat[, 1:4]  # first two trees
```

### Minimum spanning tree

`minimum_spanning_tree()` finds a minimum spanning tree using Kruskal's algorithm.
Edges are supplied in the same canonical order as `get_edges()`: pairs (u, v) with u < v, sorted by u then v.

```{r mst}
# Assign weights to the 6 edges of K4: {1,2}, {1,3}, {1,4}, {2,3}, {2,4}, {3,4}
w <- c(4, 1, 3, 2, 5, 6)
mst <- minimum_spanning_tree(k4, weights = w)
mst
```

### Sampling a random spanning tree

`sample_spanning_tree()` draws a spanning tree uniformly at random using Wilson's (1996) loop-erased random walk algorithm.

```{r sample}
set.seed(42)
sample_spanning_tree(k4)
```

### Structural analysis

Given a graph and one of its spanning trees, `fundamental_cycles()` returns the m−n+1 fundamental cycles (one per non-tree edge) and `fundamental_cuts()` returns the n−1 fundamental cuts (one per tree edge), each as a list of `adj` objects.

```{r structure}
tree <- trees[[1]]

cycles <- fundamental_cycles(k4, tree)
length(cycles)  # m - n + 1 = 6 - 4 + 1 = 3
cycles[[1]]

cuts <- fundamental_cuts(k4, tree)
length(cuts)  # n - 1 = 3
cuts[[1]]
```

### Validation

`is_spanning_tree()` checks whether a graph is structurally a spanning tree (connected, acyclic, n−1 edges).
`is_spanning_tree_of()` additionally checks that every tree edge appears in the host graph.

```{r validate}
is_spanning_tree(tree)
is_spanning_tree_of(tree, k4)
```
